---
title: "VWC prediction performance by year, summer months only"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

## Change log

- **08 Mar**: forked this from `performance_plots_byyear.Rmd`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, echo = FALSE)
```

## Contents

This notebook modifies the by-year VWC results notebook. It's identical except the input data are the correlations calculated only for the summer months (June-September).

### Basic interpretation

TBD

## Setup

Load data and packages. Set the y axis to be the same for all plots so that they are comparable. For the cross-validated values, we have a lot of negative correlations between observed and predicted so we have to set a different y-scale.

```{r, echo = TRUE}
library(ggplot2)
library(data.table)
library(purrr)

r_y <- scale_y_continuous(limits = c(0, 0.82))
r_CV_y <- scale_y_continuous(limits = c(-0.92, 0.82))

rs <- fread('data/all_rs_summer.csv')
```

```{r}
windowsFonts(`fgbook` = windowsFont("Franklin Gothic Book"))
theme_set(
  theme_bw(base_family = 'fgbook') + 
    theme(panel.grid = element_blank(),
          strip.background = element_blank(),
          axis.title.x = element_blank(), 
          legend.position = c(0.7, 0.15))
)
```


Get the properties for each of the 546 indexes. So now for each of the 546 indexes we have 5 data points: the prediction performance for each of the 5 summers. This is repeated for each of the four models: 

- `r` for just the simple regression of VWC on TWI and predicting using the model fit to all data,
- `r_depth` for the regression of VWC on TWI + depth, 
- `r_CV` for the regression of VWC on TWI with the spatially blocked cross-validation, and
- `r_depth_CV` for the spatially blocked cross-validation including both TWI + depth as predictors.

```{r}
year_x <- scale_x_continuous(breaks = 2017:2021)
```


Now we will go through the different indexes and compare model performance. In this case I will not write a bunch of musings about the patterns because there are so many. Instead I will just put them all out there for you to decide what is best.

**Again note that the y-axis is different for the cross-validation plots because there are quite a few negative observed-predicted correlation values.**

```{r define plotting function}
perf_plots <- function(rs_plot, color_factor, color_palette, color_scale_name, facet_var, plot_title) {
  leg <- guides(color = guide_legend(nrow = 3))
  col_sc <- scale_color_brewer(palette = color_palette, name = color_scale_name)
  fill_sc <- scale_fill_brewer(palette = color_palette, name = color_scale_name)
  th <- theme(axis.text.x = element_text(size = 6))
  
  p1 <- ggplot(rs_plot, aes(x = year, y = r, color = factor({{color_factor}}), fill = factor({{color_factor}}))) +
    geom_line(data = rs_plot[year > 2016]) + 
    geom_point(shape = 21, color = 'black') +
    facet_wrap(vars({{facet_var}}), labeller = label_both, scales = 'free_x') +
    year_x + r_y + leg + col_sc + fill_sc + th +
    ggtitle(plot_title, 'metric 1: VWC predicted with TWI, no CV')
  
  p2 <- ggplot(rs_plot, aes(x = year, y = r_depth, color = factor({{color_factor}}), fill = factor({{color_factor}}))) +
    geom_line(data = rs_plot[year > 2016]) + 
    geom_point(shape = 21, color = 'black') +
    facet_wrap(vars({{facet_var}}), labeller = label_both, scales = 'free_x') +
    year_x + r_y + leg + col_sc + fill_sc + th +
    ggtitle(plot_title, 'metric 2: VWC predicted with TWI + depth, no CV')
  
  p3 <- ggplot(rs_plot, aes(x = year, y = r_CV, color = factor({{color_factor}}), fill = factor({{color_factor}}))) +
    geom_line(data = rs_plot[year > 2016]) + 
    geom_point(shape = 21, color = 'black') +
    facet_wrap(vars({{facet_var}}), labeller = label_both, scales = 'free_x') +
    year_x + r_CV_y + leg + col_sc + fill_sc + th +
    ggtitle(plot_title, 'metric 3: VWC predicted with TWI, spatially blocked CV')
  
  p4 <- ggplot(rs_plot, aes(x = year, y = r_depth_CV, color = factor({{color_factor}}), fill = factor({{color_factor}}))) +
    geom_line(data = rs_plot[year > 2016]) + 
    geom_point(shape = 21, color = 'black') +
    facet_wrap(vars({{facet_var}}), labeller = label_both, scales = 'free_x') +
    year_x + r_CV_y + leg + col_sc + fill_sc + th +
    ggtitle(plot_title, 'metric 4: VWC predicted with TWI + depth, spatially blocked CV')
  
  walk(list(p1, p2, p3, p4), print)
}

```


## SWI 

### No filter

For each of the four prediction metrics, within each DEM resolution, suction factor varies.

```{r}
perf_plots(rs[index=='SWI' & filter=='NFNW'], color_factor = suction_factor, color_palette = 'Reds', color_scale_name = 'suction factor', facet_var = dem_resolution, plot_title = 'SWI, no filter') 
```

### Gaussian filter

For each of the four prediction metrics, within each sigma, suction factor varies.

```{r}
perf_plots(rs[index=='SWI' & filter=='gauss'], color_factor = suction_factor, color_palette = 'Reds', color_scale_name = 'suction factor', facet_var = sigma, plot_title = 'SWI, Gaussian filter') 
```

### SSC filter

For each of the four prediction metrics, within each filter window, suction factor varies.

```{r}
perf_plots(rs[index=='SWI' & filter=='SSC'], color_factor = suction_factor, color_palette = 'Reds', color_scale_name = 'suction factor', facet_var = filter_window, plot_title = 'SWI, SSC filter') 
```

## TWI

Here we want to distinguish between the performance of the different flow algorithms. There are 9 flow algorithms. Most of them do not have any additional differentiation in input parameters but the TWI calculations with the MFD and MTFD algorithms were done with 6 different convergence factors each. So we basically have 19 different flow algorithms. To prevent the plots from becoming too busy, in each case, I will plot the 7 flow algorithms without any different parameters on one set of panels, then the MFD by convergence factor, then the MTFD by convergence factor.

### No filter

#### All flow algorithms besides MFD and MTFD

Within TWI index by flow algorithm when no filter is used, DEM resolution varies.

```{r}
perf_plots(rs[index=='TWI' & filter=='NFNW' & !flow_algorithm %in% c('MFD', 'MTFD')], color_factor = flow_algorithm, color_palette = 'Dark2', color_scale_name = 'flow algorithm', facet_var = dem_resolution, plot_title = 'TWI, no filter, 7 of 9 flow algorithms') 
```

#### MFD

Within the flow algorithm MFD with no filter, first DEM resolution varies, then convergence factor varies.

```{r}
perf_plots(rs[index=='TWI' & filter=='NFNW' & flow_algorithm %in% c('MFD')], color_factor = convergence_factor, color_palette = 'Blues', color_scale_name = 'convergence factor', facet_var = dem_resolution, plot_title = 'TWI, no filter, MFD algorithm by convergence factor') 
```

#### MTFD

Within the flow algorithm MTFD with no filter, first DEM resolution varies, then convergence factor varies.

```{r}
perf_plots(rs[index=='TWI' & filter=='NFNW' & flow_algorithm %in% c('MTFD')], color_factor = convergence_factor, color_palette = 'Blues', color_scale_name = 'convergence factor', facet_var = dem_resolution, plot_title = 'TWI, no filter, MTFD algorithm by convergence factor') 
```

### Gaussian filter

#### All flow algorithms besides MFD and MTFD

Within TWI index by flow algorithm with Gaussian filter, sigma varies.

```{r}
perf_plots(rs[index=='TWI' & filter=='gauss' & !flow_algorithm %in% c('MFD', 'MTFD')], color_factor = flow_algorithm, color_palette = 'Dark2', color_scale_name = 'flow algorithm', facet_var = sigma, plot_title = 'TWI, Gaussian filter, 7 of 9 flow algorithms') 
```

#### MFD

Within the flow algorithm MFD with Gaussian filter, first sigma varies, then convergence factor varies.

```{r}
perf_plots(rs[index=='TWI' & filter=='gauss' & flow_algorithm %in% c('MFD')], color_factor = convergence_factor, color_palette = 'Blues', color_scale_name = 'convergence factor', facet_var = sigma, plot_title = 'TWI, Gaussian filter, MFD algorithm by convergence factor') 
```

#### MTFD

Within the flow algorithm MTFD with Gaussian filter, first sigma varies, then convergence factor varies.

```{r}
perf_plots(rs[index=='TWI' & filter=='gauss' & flow_algorithm %in% c('MTFD')], color_factor = convergence_factor, color_palette = 'Blues', color_scale_name = 'convergence factor', facet_var = sigma, plot_title = 'TWI, Gaussian filter, MTFD algorithm by convergence factor') 
```


### SSC filter

#### All flow algorithms besides MFD and MTFD

Within TWI index by flow algorithm with SSC filter, filter window varies.

```{r}
perf_plots(rs[index=='TWI' & filter=='SSC' & !flow_algorithm %in% c('MFD', 'MTFD')], color_factor = flow_algorithm, color_palette = 'Dark2', color_scale_name = 'flow algorithm', facet_var = filter_window, plot_title = 'TWI, Gaussian filter, 7 of 9 flow algorithms') 
```

#### MFD

Within the flow algorithm MFD with SSC filter, first filter window varies, then convergence factor varies.

```{r}
perf_plots(rs[index=='TWI' & filter=='SSC' & flow_algorithm %in% c('MFD')], color_factor = convergence_factor, color_palette = 'Blues', color_scale_name = 'convergence factor', facet_var = filter_window, plot_title = 'TWI, Gaussian filter, MFD algorithm by convergence factor') 
```

#### MTFD

Within the flow algorithm MTFD with SSC filter, first filter window varies, then convergence factor varies.

```{r}
perf_plots(rs[index=='TWI' & filter=='SSC' & flow_algorithm %in% c('MTFD')], color_factor = convergence_factor, color_palette = 'Blues', color_scale_name = 'convergence factor', facet_var = filter_window, plot_title = 'TWI, Gaussian filter, MTFD algorithm by convergence factor') 
```
