---
title: "2017-2021 seasonal correlations by day"
author: "Quentin D. Read"
date: "8/31/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## change log

- 8/30/22: This script was copied and modified from `alldryperiodsbyday.Rmd`.
- 8/31/22: Change months to Edwin's new specification; add boxplots

This document has the following analysis:

- Get VWC averages by single day (no moving window) for the months identified as recharge (Nov), saturation (Apr), and use (Sep) periods.
- Correlate these with TWI, using only the highest performing algorithm(s), for each day in the period.
- Plot the results.

Read data.

```{r}
library(data.table)
library(purrr)
library(ggplot2)
library(lubridate)

# Read in raw data
moisture_qc <- fread('C:/Users/qdread/onedrive_usda/ars_projects/libohova/data/processed_moisture_qc.csv')
twi <- fread('C:/Users/qdread/onedrive_usda/ars_projects/libohova/data/processed_twi.csv')
logger_lookup <- fread('data/LoggerLocations.csv')

# Specify dry periods
dry_periods <- data.table(type = rep(c('recharge', 'saturation', 'use'), each = 5),
                          month = rep(c(11, 4, 9), each = 5),
                          year = 2017:2021)

# port should be character variable
moisture_qc[, port := as.character(port)]
twi[, port := as.character(port)]
```

Get daily VWC averages for each day of each period. To do this we need to join the period data frame, extended to have a row for each day, to the VWC data frame, thereby assigning each day to one (or none) of the periods.

```{r}
dry_periods[, start := as.Date(paste(year, month, '01', sep = '-'))]
dry_periods[, end := as.Date(paste(year, month, days_in_month(start), sep = '-'))]
dry_periods[, period_ID := 1:.N]

dry_days <- dry_periods[, .(date = seq(from = start, to = end, by = 1),
                            day_from_start = c(0, seq(as.numeric(end - start)))), by = .(period_ID, type)]

moisture_qc <- moisture_qc[dry_days, on = .(date)]
vwc <- moisture_qc[, .(VWC = mean(VWC_QC)), by = .(period_ID, type, year, date, day_from_start, logger, port, depth)]
```

Plot them to generally see the VWC trend over each period and by soil depth. Classify this by type of period.

```{r}
datexdepth <- vwc[, .(VWC = mean(VWC)), by = .(period_ID, type, year, day_from_start, depth)]

ggplot(datexdepth, aes(x = day_from_start, color = factor(depth), group = interaction(depth, period_ID), y = VWC)) +
  facet_wrap(~ type) +
  geom_point() + geom_line() + theme_minimal()
```

I am going to use TWI index calculated from the FD8 flow algorithm at a DEM resolution of 7 m, unfiltered, to do this analysis because if you average prediction performance by month, it has the highest *cross-validated* performance of all indexes regardless of whether or not depth was included as a covariate. Of course you can always change this. I doubt it will have an enormous influence on this result.

Get the index in question and join it with VWC.

```{r}
idx_use <- 'TWI.FD8.res7.NFNW'
twi_use <- twi[index_raw %in% idx_use, .(logger, port, TWI)]
vwc[twi_use, on = .(logger, port), TWI := i.TWI]
```

Do the correlation for each day. (Not cross-validated, just the observed-predicted correlation). **Note**: We do not have any VWC date from before 4/14/2017 so the correlation for April 2017 is based on only part of the month.

```{r}
get_corr <- function(VWC, TWI, depth, use_depth) {
  if (!use_depth) fit <- lm(VWC ~ TWI) else fit <- lm(VWC ~ TWI + depth)
  pred <- predict(fit)
  cor(VWC, pred)
}

corrs_time <- vwc[date >= "2017-04-14", .(r = get_corr(VWC, TWI, depth, use_depth = FALSE), r_depth = get_corr(VWC, TWI, depth, use_depth = TRUE)), by = .(period_ID, type, year, date, day_from_start)]
```

Plot them.

```{r}
type_facet <- facet_wrap(~ type, labeller = labeller(type = c(recharge = 'recharge (November)', saturation = 'saturation (April)', use = 'use (September)')))

ggplot(corrs_time, aes(x = day_from_start, color = factor(year))) +
  type_facet +
  geom_point(aes(y = r)) +
  geom_line(aes(y = r, group = period_ID)) +
  theme_minimal() +
  scale_color_brewer(name = 'year', palette = 'Dark2') +
  ggtitle('ignoring soil depth') +
  theme(legend.position = 'bottom')

ggplot(corrs_time, aes(x = day_from_start, color = factor(year))) +
  type_facet +
  geom_point(aes(y = r_depth)) +
  geom_line(aes(y = r_depth, group = period_ID)) +
  theme_minimal() +
  scale_color_brewer(name = 'year', palette = 'Dark2') +
  ggtitle('accounting for soil depth') +
  theme(legend.position = 'bottom')
```

## Boxplots of VWC for each month

This plot lumps all soil depths together. Again, April 2017 is based on only half the month.

```{r}
ggplot(vwc, aes(y = VWC, x = year, fill = factor(year))) +
  type_facet +
  geom_boxplot() +
  theme_minimal() +
  scale_fill_brewer(name = 'year', palette = 'Dark2') +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = 'none')
```

## Boxplots of correlations for each month

Separate figure made for the correlations that account for soil depth and those that ignore it. As before, April 2017 is based on only half the month.

```{r}
ggplot(corrs_time, aes(y = r, x = year, fill = factor(year))) +
  type_facet +
  geom_boxplot() +
  theme_minimal() +
  scale_fill_brewer(name = 'year', palette = 'Dark2') +
  theme(legend.position = 'none') +
  ggtitle('ignoring soil depth')

ggplot(corrs_time, aes(y = r_depth, x = year, fill = factor(year))) +
  type_facet +
  geom_boxplot() +
  theme_minimal() +
  scale_fill_brewer(name = 'year', palette = 'Dark2') +
  theme(legend.position = 'none') +
  ggtitle('accounting for soil depth')
```

## Boxplot figures formatted for manuscript

```{r}
left_panel <- ggplot(vwc, aes(y = VWC, x = year, fill = factor(year))) +
  type_facet +
  geom_boxplot() +
  theme_minimal() +
  scale_fill_brewer(name = 'year', palette = 'Dark2') +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = 'none')

right_panel <- ggplot(corrs_time, aes(y = r_depth, x = year, fill = factor(year))) +
  type_facet +
  geom_boxplot() +
  theme_minimal() +
  scale_fill_brewer(name = 'year', palette = 'Dark2') +
  scale_y_continuous(name = 'Performance of models incorporating depth (r)') +
  theme(legend.position = 'none') 

library(gridExtra)
png('C:/Users/qdread/onedrive_usda/ars_projects/libohova/figures/formatted_seasonal_figure.png', height = 4, width = 11, res = 400, units = 'in')
grid.arrange(left_panel, right_panel, nrow = 1)
dev.off()
```

